Effectiveness of community-based marine reserves in small-scale fisheries

### 2018-09-24 docs/paper/Villasenor-Derbez.Rmd
Finalized manuscript. Waiting for Ere's adscription

---
title: "Effectiveness of community-based marine reserves in small-scale fisheries"
rtitle: "Community-based marine reserves"
author: "Juan Carlos Villaseñor-Derbez$^{1,2,*}$, Eréndira Aceves-Bueno$^{1}$, Álvin Suarez$^{2}$, Stuart Fulton$^{2}$, Arturo Hernández-Velasco$^{2}$, Jorge Torre$^{2}$, Fiorenza Micheli$^{3}$"
institution: "$^{1}$Bren School of Environmental Science and Management, University of California, Santa Barbara, Santa Barbara, CA, USA\\newline
$^{2}$Comunidad y Biodiversidad A.C., Guaymas, Sonora, Mexico\\newline
$^{3}$Hopkins Marine Station and Center for Ocean Solutions, Stanford University, Pacific Grove, CA, USA"
corrauth: "Juan Carlos Villaseñor-Derbez, Bren Hall, University of California, Santa Barbara, Santa Barbara, CA, 93106"
email: "jvillasenor@bren.ucsb.edu"
abstract: "Coastal marine ecosystems provide livelihoods for small-scale fishers and coastal communities around the world. Artisanal fisheries face great challenges since they are difficult to monitor, enforce, and manage. Combining territorial user rights for fisheries (TURF) with no-take marine reserves to create TURF-reserves is believed to improve the performance of small-scale fisheries by buffering fisheries from environmental variability and management errors, while ensuring that fishers reap the benefits of conservation investments. In the last six years, and following a 2012 regulation, 18 TURF-reserves have been implemented in Mexico; their effectiveness has not been formally evaluated. We combine causal inference techniques and a social-ecological systems framework to provide a holistic evaluation of community-based TURF reserves in three coastal communities in Mexico. We find that while reserves have not yet achieved their stated goal of increasing lobster densities, they continue to receive significant support from the fishing communities. A lack of ecological and socioeconomic effects likely results from a combination of factors. First, the lobster fisheries are already well managed, and it is unlikely that reserves might have a detectable effect. Second, some of the reserves are not large enough to protect lobsters' home ranges. Third, some of these reserves might be too young for the effects to show. However, these reserves have shaped small-scale fishers’ way of thinking about marine conservation, which can provide a foundation for establishing additional, larger marine reserves needed to effectively conserve mobile species."
keywords: "TURF-reserves, Causal Inference, Social-Ecological Systems, Marine Protected Areas, Marine Conservation, Small-Scale Fisheries"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    template: template.tex
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

@@

```{r, fig.cap = "\\label{fig:map}Location of the three coastal communities studied (A). Isla Natividad (B) is located off the Baja California Peninsula, Maria Elena and Punta Herrero (C) are located in the Yucatan Peninsula. Blue polygons represent the TURFs, and red polygons the marine reserves."}
# Define reference map

@@

plot_grid(left, ph, labels = c("", "C"), ncol = 2)
```

@@

Where year-fixed effects are represented by $\gamma_{it} Year_t$, and $\beta Zone_i$ captures the difference between reserve ($Zone = 1$) and control ($Zone = 0$) sites. The interaction term $\lambda_{it} Year_t\times Zone_i$ represents the mean change in the indicator inside the reserve, for year $t$, with respect to the year of implementation in the control site (See Table \ref{table:com_sum}). When evaluating biomass and densities of the invertebrate or fish communities, we include $\sigma_j$ to control for species-fixed effects.

@@

We use the social-ecologycal system (SES) framework to evaluate each community as a means of providing an explanation to the biological and socioeconomic results. The use the SES framework standardizes our analysis and allows us to communicate our results in a common language across fields. We based our variable selection primarily on \citet{leslie_2015-na,basurto_2013-oq}, who have previously analyzed Mexican fishing cooperatives using this framework. We also incorporate other relevant information following \citet{difranco_2016-Xw,edgar_2014-UO} Table \ref{table:ses} shows the selected variables, their definition and selected indicators.

```{r}
read.csv(here::here("raw_data", "ses.csv")) %>% 
  select(-Dim) %>% 
  knitr::kable(caption = "\\label{table:ses}Variables used in our Social-Ecological System analysis (IN = Isla Natividad, ME = Maria Elena, PH = Punta Herrero).",
               format = "latex",
               escape = F) %>% 
  kable_styling(latex_options = "HOLD_position") %>% 
  column_spec(column = 1, width = "6cm") %>% 
  column_spec(column = 2, width = "12cm") %>% 
  group_rows(group_label = "Resource System", start_row = 1, end_row = 4) %>% 
  group_rows(group_label = "Resource Unit", start_row = 5, end_row = 6) %>% 
  group_rows(group_label = "Actors", start_row = 7, end_row = 8) %>% 
  group_rows(group_label = "Governance system", start_row = 9, end_row = 12)
```

@@

Lobster landings and revenue were only available for Isla Natividad and Maria Elena (Fig \ref{fig:lobsters}). For all years before implementation, the effect sizes are close to zero, indicating that the control and treatment sites have similar pre-treatment trends, suggesting that these are plausible controls. However, effect sizes do not change after the implementation of the reserve. Interestingly, the negative effect observed for Isla Natividad on year 5 correspond to the 2011 hypoxia events. The only positive change observed in lobster landings is for Isla Natividad in 2014 ($p < 0.1$). The three years of post-implementation data for Maria Elena do not show a significant effect of the reserve. Isla Natividad shows higher revenues after the implementation of the reserve, as compared to the control communities. However, these changes are not significant and are associated to increased variation. Full tables with model coefficients are presented in the supplementary materials (S4 Table, S5 Table.

<!-- Revenues show a positive trend post implementation for I Natividad. Because landings have not changed, this might be due to an increase in price? -->

@@

Although we have little information on the social dimension of these fisheries, we can use the social-ecological systems framework to analyze the performance of each governance system (Table \ref{tab:ses}). Our analysis shows that all communities analyzed share similarities in their Governance system which is based on cooperatives (GS5.2.3.2), with strong rules in use that include Operational rules (GS6.2), Collective-choice rules (GS6.3), Constitutional rules (GS6.3), and even Territorial use communal rights (GS6.1.4.3). However, we identified important differences in terms of the actors, resource systems, and resource units. Although all communities show a high level of leadership (A5), the level of trust (A6.1) is lower in Punta Herrero. In general, the presence and success of conservation initiatives depends on the incentives of local communities to maintain a healthy status of the resources they depend upon \citep{jupiter_2017}. The enabling conditions for conservation seem to be strongly present in all communities. Due to the clarity of access rights and isolation, the benefits of conservation directly benefit the members of the fishing cooperative. These conditions have favored the development of efficient community-based enforcement systems.

```{r, fig.width = 6.5, fig.height = 4, fig.cap = "\\label{fig:indicators}Effect sizes for marine reserves from Isla Natividad (IN; red cirlcles), Maria Elena (ME; blue triangles), and Punta Herrero (PH; green squares) for lobster densities (*Panulirus spp*; A), fish biomass (B), invertebrate densities (C), and fish densities (D). Plots are ordered by survey type (left column: invertebrates; right column: fish). Points are jittered hotizontally to avoid overplotting. Points indicate the effect size and standard errors. Years have been centered to year of implementation."}

@@

plot2 <- cowplot::plot_grid(plotlist = list(L_plot, F_B, I_N, F_N),
                            labels = "AUTO",
                            ncol = 2)

plot2
```

@@

```{r, fig.width = 5, fig.height = 4, fig.cap = "\\label{fig:lobsters}Effect sizes for lobster catches (A) and revenues (B) in at Isla Natividad (IN; red circles) and Maria Elena (ME; blue triangles). Points indicate the effect size and standard errors. Years have been centered to year of implementation."}

@@

plot_grid(c_plot, v_plot, ncol = 1, labels = "AUTO")

plot3 <- plot_grid(c_plot + theme(legend.position = "none"), v_plot, ncol = 2, labels = "AUTO")

@@

Our results indicate that these TURF-reserves have not increased lobster densities. Additionally, no co-benefits were identified when using other ecological indicators other than the previously reported buffering effect that reserves can have to environmental variability in Isla Natividad \citep{micheli_2012-EU}. The socioeconomic indicators pertaining landings and revenues showed little to no change after reserve implementation. The coastal ecosystems where these reserves are located have been profoundly affected by climatic and oceanographic extremes, including warming events, extreme storms and prolonged hypoxia \citep{micheli_2012-EU}, (**Beas et al. in prep., Woodson et al in review**). Despite the lack of evidence of the effectiveness of these reserves, most of the communities show a positive perception about their performance and continue to support their presence \citep{ayer_2018}. Understanding the social-ecological context in which these communities and their reserves operate might provide insights to this.

@@

While the evaluated reserves have failed to provide fishery benefits up to now, there are a number of additional ecological, fisheries, and social benefits. Marine reserves provide protection to a wider range of species and vulnerable habitat, like coral reefs. These sites can serve as an insurance against uncertainty and errors in fisheries management, as well as environmental shocks \citep{hilborn_2004,hilborn_2006,micheli_2012-EU} (**,Aalto et al, in press**). Self-regulation of fishing effort (*i.e.* reduction in harvest) can serve as a way to compensate for future declines associated to environmental variation \citep{finkbeiner_2018}. Embarking in a marine conservation project can bring the community together, which promotes social cohesion and builds social capital \citep{fulton_2019}. Furthermore, showing commitment to marine conservation and sustainable fishing practices allows fishers to have greater bargaining power and leverage over fisheries management \citep{prezramrez_2012}. Furthermore, the lack of effectiveness observed in these reserves may not be generalizable to other reserves established under the same legal frameowrk (*i.e.* Fish Refges) in Mexico, and future research should aim at evaluating other areas that have also been established as bottom-up processes bt without the presence of TURFs (*e.g.* \citet{dof_websiteC_2012}), or others established through a top-down process (*i.e.* \citet{dof_websiteU_2018}).

@@

>>>>

---
title: "Effectiveness of community-based marine reserves in small-scale fisheries"
rtitle: "Community-based marine reserves"
author: "Juan Carlos Villaseñor-Derbez$^{1,2,*}$, Eréndira Aceves-Bueno$^{1}$, Álvin Suarez$^{2}$, Stuart Fulton$^{2}$, Arturo Hernández-Velasco$^{2}$, Jorge Torre$^{2}$, Fiorenza Micheli$^{3}$"
institution: "$^{1}$Bren School of Environmental Science and Management, University of California, Santa Barbara, Santa Barbara, CA, USA\\newline
$^{2}$Comunidad y Biodiversidad A.C., Guaymas, Sonora, Mexico\\newline
$^{3}$Hopkins Marine Station and Center for Ocean Solutions, Stanford University, Pacific Grove, CA, USA"
corrauth: "Juan Carlos Villaseñor-Derbez, Bren Hall, University of California, Santa Barbara, Santa Barbara, CA, 93106"
email: "juancarlos@ucsb.edu"
abstract: "Coastal marine ecosystems provide livelihoods for small-scale fishers and coastal communities around the world. Artisanal fisheries face great challenges since they are difficult to monitor, enforce, and manage. Combining territorial user rights for fisheries (TURF) with no-take marine reserves to create TURF-reserves is believed to improve the performance of small-scale fisheries by buffering fisheries from environmental variability and management errors, while ensuring that fishers reap the benefits of conservation investments. In the last six years, and following a 2012 regulation, 18 TURF-reserves have been implemented in Mexico; their effectiveness has not been formally evaluated. We combine causal inference techniques and a social-ecological systems framework to provide a holistic evaluation of community-based TURF reserves in three coastal communities in Mexico. We find that while reserves have not yet achieved their stated goal of increasing lobster densities, they continue to receive significant support from the fishing communities. A lack of ecological and socioeconomic effects likely results from a combination of factors. First, the lobster fisheries are already well managed, and it is unlikely that reserves might have a detectable effect. Second, some of the reserves are not large enough to protect lobsters' home ranges. Third, some of these reserves might be too young for the effects to show. However, these reserves have shaped small-scale fishers’ way of thinking about marine conservation, which can provide a foundation for establishing additional, larger marine reserves needed to effectively conserve mobile species."
keywords: "TURF-reserves, Causal Inference, Social-Ecological Systems, Marine Protected Areas, Marine Conservation, Small-Scale Fisheries"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    template: template.tex
---

@@

```{r}
# Define reference map

@@

plot1 <- plot_grid(left, ph, labels = c("", "C"), ncol = 2)
```

@@

Where year-fixed effects are represented by $\gamma_{it} Year_t$, and $\beta Zone_i$ captures the difference between reserve ($Zone = 1$) and control ($Zone = 0$) sites. The interaction term $\lambda_{it} Year_t\times Zone_i$ represents the mean change in the indicator inside the reserve, for year $t$, with respect to the year of implementation in the control site. When evaluating biomass and densities of the invertebrate or fish communities, we include $\sigma_j$ to control for species-fixed effects.

@@

We use the social-ecologycal system (SES) framework to evaluate each community as a means of providing an explanation to the biological and socioeconomic results. The use the SES framework standardizes our analysis and allows us to communicate our results in a common language across fields. We based our variable selection primarily on \citet{leslie_2015-na,basurto_2013-oq}, who have previously analyzed Mexican fishing cooperatives using this framework. We also incorporate other relevant information following \citet{difranco_2016-Xw} and \citet{edgar_2014-UO}. Table \ref{table:ses} shows the selected variables, their definition and selected indicators.

@@

Lobster landings and revenue were only available for Isla Natividad and Maria Elena (Fig \ref{fig:lobsters}). For all years before implementation, the effect sizes are close to zero, indicating that the control and treatment sites have similar pre-treatment trends, suggesting that these are plausible controls. However, effect sizes do not change after the implementation of the reserve. Interestingly, the negative effect observed for Isla Natividad on year 5 correspond to the 2011 hypoxia events. The only positive change observed in lobster landings is for Isla Natividad in 2014 ($p < 0.1$). The three years of post-implementation data for Maria Elena do not show a significant effect of the reserve. Isla Natividad shows higher revenues after the implementation of the reserve, as compared to the control communities. However, these changes are not significant and are associated to increased variation. Full tables with model coefficients are presented in the supplementary materials (S4 Table, S5 Table).

@@

Our analysis of the SES (Table \ref{table:ses}) shows that all analyzed communities share similarities known to foster sustainable resource management and increase reserve effectiveness. For example, fishers opperate within clearly outlined TURFs (RS2, GS6.1.4.3) that provide exclusive access to resources and reserves. Along with their relatively small groups (A1 - Number of relevant actors), Isolation (A3), Operational rules (GS6.2), Social monitoring (GS9.1), and Graduated sanctions (GS10.1), these fisheries have solid governance structures that enable them to monitor their resources and enforce rules to ensure sustainable management. In general, success of conservation initiatives depends on the incentives of local communities to maintain a healthy status of the resources upon which they depend \citep{jupiter_2017}. Due to the clarity of access rights and isolation, the benefits of conservation directly benefit the members of the fishing cooperatives, which have favored the development of efficient community-based enforcement systems. However, our SES analysis also highlights factors that might hinder reserve performance or mask outcomes While total reserve size ranges from 0.2% to 3.7% of the TURF area, individual reserves are often small (RS3), and relatively young (RS5). Additionally, fishers harvest healthy stocks (RS4.1), and it's unlikely that marine reserves will result in increased catches.

```{r}

@@

plot2 <- cowplot::plot_grid(plotlist = list(L_plot, F_B, I_N, F_N),
                            labels = "AUTO",
                            ncol = 2)
```

@@

```{r}

@@

plot3 <- plot_grid(c_plot, v_plot, ncol = 1, labels = "AUTO")
```

@@

Our results indicate that these TURF-reserves have not increased lobster densities. Additionally, no co-benefits were identified when using other ecological indicators other than the previously reported buffering effect that reserves can have to environmental variability in Isla Natividad \citep{micheli_2012-EU}. The socioeconomic indicators pertaining landings and revenues showed little to no change after reserve implementation. The coastal ecosystems where these reserves are located have been profoundly affected by climatic and oceanographic extremes, including warming events, extreme storms and prolonged hypoxia \citep{micheli_2012-EU,woodson}. Despite the lack of evidence of the effectiveness of these reserves, most of the communities show a positive perception about their performance and continue to support their presence \citep{ayer_2018}. Understanding the social-ecological context in which these communities and their reserves operate might provide insights to this.

@@

While the evaluated reserves have failed to provide fishery benefits up to now, there are a number of additional ecological, fisheries, and social benefits. Marine reserves provide protection to a wider range of species and vulnerable habitat, like coral reefs. These sites can serve as an insurance against uncertainty and errors in fisheries management, as well as environmental shocks \citep{hilborn_2004,hilborn_2006,micheli_2012-EU,aalto}. Self-regulation of fishing effort (*i.e.* reduction in harvest) can serve as a way to compensate for future declines associated to environmental variation \citep{finkbeiner_2018}. Embarking in a marine conservation project can bring the community together, which promotes social cohesion and builds social capital \citep{fulton_2019}. Showing commitment to marine conservation and sustainable fishing practices allows fishers to have greater bargaining power and leverage over fisheries management \citep{prezramrez_2012}. Furthermore, the lack of effectiveness observed in these reserves may not be generalizable to other reserves established under the same legal frameowrk (*i.e.* Fish Refges) in Mexico, and future research should aim at evaluating other areas that have also been established as bottom-up processes bt without the presence of TURFs (*e.g.* \citet{dof_websiteC_2012}), or others established through a top-down process (*i.e.* \citet{dof_websiteU_2018}).

@@

```{r, fig.cap = "\\label{fig:map}Location of the three coastal communities studied (A). Isla Natividad (B) is located off the Baja California Peninsula, Maria Elena and Punta Herrero (C) are located in the Yucatan Peninsula. Blue polygons represent the TURFs, and red polygons the marine reserves."}
plot1
```

```{r, fig.width = 6.5, fig.height = 4, fig.cap = "\\label{fig:indicators}Effect sizes for marine reserves from Isla Natividad (IN; red cirlcles), Maria Elena (ME; blue triangles), and Punta Herrero (PH; green squares) for lobster densities (*Panulirus spp*; A), fish biomass (B), invertebrate densities (C), and fish densities (D). Plots are ordered by survey type (left column: invertebrates; right column: fish). Points are jittered hotizontally to avoid overplotting. Points indicate the effect size and standard errors. Years have been centered to year of implementation."}
plot2
```

```{r, fig.width = 5, fig.height = 4, fig.cap = "\\label{fig:lobsters}Effect sizes for lobster catches (A) and revenues (B) in at Isla Natividad (IN; red circles) and Maria Elena (ME; blue triangles). Points indicate the effect size and standard errors. Years have been centered to year of implementation."}
plot3
```

```{r}
read.csv(here::here("raw_data", "ses.csv")) %>% 
  select(-Dim) %>% 
  knitr::kable(caption = "\\label{table:ses}Variables for the Social-Ecological System analysis (IN = Isla Natividad, ME = Maria Elena, PH = Punta Herrero). Alphanumeric codes follow \\citet{basurto_2013-oq}; an asterisk (*) denotes variables incorporated based on \\citet{difranco_2016-Xw} and \\citet{edgar_2014-UO}.",
               format = "latex",
               escape = F) %>% 
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size = 11) %>% 
  column_spec(column = 1, width = "6.5cm") %>% 
  column_spec(column = 2, width = "12cm") %>% 
  group_rows(group_label = "Resource System (RS)", start_row = 1, end_row = 5) %>% 
  group_rows(group_label = "Resource Unit (RU)", start_row = 6, end_row = 6) %>% 
  group_rows(group_label = "Actors (A)", start_row = 7, end_row = 8) %>% 
  group_rows(group_label = "Governance system (G)", start_row = 9, end_row = 12)
```
